\thetitle

### 2008-01-17 paper.tex
Minor edits.

The interaction of wireless user and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
\footnote{This is violated by some quality of service (QoS) schemes. However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid. The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.

@@

In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications. This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks. Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.  Experimental deployment sidesteps the issue of accurately modeling the lower layers of the network. Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.

@@

We do not assume or claim that the traffic found at \caps{IETF60} is representative of conference settings in general. The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting. We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases. Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace. In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area. During working group sessions, we see highly concentrated, heavy usage patterns. At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment. At the nadir, a lone node sent only a single 61-byte packet in 10 minutes. All levels of activity between these extremes are represented. Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior. This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation. The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable. %For a traffic model to be truly realistic, it must be realistic across many different usage cases.

@@

We use the Qualnet wireless network simulator to perform our experiments. We simulate a stationary multi-hop 802.11b network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol, with nodes placed randomly in a square field with sides of 1500 meters.
In addition to the active nodes corresponding to trace \caps{IP}s, equally many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

@@

We separate our traffic generation models into three orthogonal and nearly independent levels of behavior:
\begin{enumerate}
\item \textbf{Flow End-Point Topology}: which nodes communicate with each other, and how frequently; i.e. how flow end-points are mapped onto nodes in the network.
\item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
\item \textbf{Packet Behavior}: sizes of individual packets, and the intervals between their transmission.
\end{enumerate}
For each level of traffic behavior, we compare several different behavior models. The different levels of models and variants at each level are listed and described in Table~\ref{tab:traffic-models} and illustrated with specific examples in Figure~\ref{fig:traffic-models}.

@@

We have selected nine performance metrics at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), \caps{RREQ}s initiated per node, routing queue drop rate.
\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate, retransmission failure rate.
\end{enumerate}
These metrics are commonly used to evaluate the performance of new wireless protocols. For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

@@

\vfill

>>>>

The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
\footnote{This is violated by some quality of service (QoS) schemes. However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid. The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.

@@

In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications. This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks. Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.  Experimental deployments sidestep the problem of accurately modeling the lower layers of the network. Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.

@@

We do not assume or claim that the traffic found at \caps{IETF60} is representative of conference settings in general. The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting. We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases. Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace. In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area. During working group sessions, we see highly concentrated, heavy usage patterns. At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment. At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes. All levels of activity between these extremes are represented. Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior. This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation. The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable. %For a traffic model to be truly realistic, it must be realistic across many different usage cases.

@@

We use the Qualnet wireless network simulator to perform our experiments. We simulate a stationary multi-hop 802.11b network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

@@

We separate our traffic generation models into three orthogonal and nearly independent levels of behavior:
\begin{enumerate}
\item \textbf{Network Behavior}: which nodes communicate with each other, and how frequently; \textit{i.e.} how flow end-points are mapped onto nodes in the network.
\item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
\item \textbf{Packet Behavior}: sizes of individual packets, and the intervals between their transmission.
\end{enumerate}
For each level of traffic behavior, we compare several different behavior models. The different levels of models and variants at each level are listed and described in Table~\ref{tab:traffic-models} and illustrated with specific examples in Figure~\ref{fig:traffic-models}.

% TODO: more here on models...
% TODO: rewrite next paragraph too...

@@

We have selected a variety of performance metrics at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), routing queue drop rate.
\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate.
% TODO: are we actually using RTS/CTS?
\end{enumerate}
These metrics are commonly used to evaluate the performance of new wireless protocols. For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

@@

%\vfill
